\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ...
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex
\usepackage{amssymb}
\usepackage{url}
\usepackage{stmaryrd}

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

\title{Assignment 4}

\begin{document}
\maketitle

\section*{$k$-induction}
In class and notes,\footnote{\url{https://github.com/barghouthi/cs704/blob/master/notes/transRelEnc.pdf}}
 we discussed $k$-induction as a technique
for proving validity of a Hoare triple.
You are to implement the $k$-induction algorithm
using the Z3 SMT solver.


Your implementation should take a Hoare triple:
$$\{\phi\}\ \texttt{while } b \texttt{ do } P_\emph{body}\  \{\psi\}$$
Given a value $k \geq 1$,
your implementation should perform
$k$-induction, returning ``success'' if
$k$-induction succeeds, and ``failure'' otherwise.

\emph{Your implementation can take the program
as two formulas}, $b$ and $\texttt{enc}(P_{\emph{body}})$.
So you do not have to parse the program and encode automatically
for this assignment.

\begin{itemize}
  \item It is highly recommend that you use the Python
  API of the Z3 SMT solver (\url{github.com/Z3Prover/z3}).
  Z3 has other language bindings, but Python is the easiest
  to prototype with and will make you most productive.

\item You are expected to \textbf{exhaustively comment your
code} and describe what parts of $k$-induction you are implementing.

\item You should encode three non-trivial Hoare triples.
All examples you provide should be valid Hoare triples.
One of the Hoare triples should require $k\geq2$
for the induction to succeed.
Explain the programs you provide.
For the example where $k\geq 2$ is required, explain why
$k=1$ is insufficient.
It is OK for your examples to be hard-coded.

\item Provide a valid Hoare triple
for which $k$-induction cannot prove correctness
for any value of $k$. Argue why that is the case.

\item Note: In the notes, we encoded sequential 
    composition with an existential quantifier.
    The quantifiers are not needed when 
    checking validity, as is done in $k$-induction,
    so you do not need them.
\end{itemize}


\end{document}
